---
title: "Introduction to Statistical Learning"
author: "YOUR NAME"
output:
  pdf_document: default
editor_options:
  chunk_output_type: console
---

**1. For each of parts (a) through (d), indicate whether we would generally expect the performance of a flexible statistical learning method to be better or worse than an inflexible method. Justify your answer.**

(a) The sample size n is extremely large, and the number of predictors p is small.

(b) The number of predictors p is extremely large, and the number of observations n is small.

(c) The relationship between the predictors and response is highly non-linear. 

(d) The variance of the error terms, i.e. $\sigma^2 = Var(\epsilon)$, is extremely high.


2. Explain whether each scenario is a classification or regression problem, and indicate whether we are most interested in inference or prediction. Finally, provide n and p.

(a) We collect a set of data on the top 500 firms in the US. For each firm we record profit, number of employees, industry and the
CEO salary. We are interested in understanding which factors affect CEO salary.

(b) We are considering launching a new product and wish to know whether it will be a success or a failure. We collect data on 20
similar products that were previously launched. For each product we have recorded whether it was a success or failure, price charged for the product, marketing budget, competition price, and ten other variables.

(c) We are interested in predicting the % change in the USD/Euro exchange rate in relation to the weekly changes in the world
stock markets. Hence we collect weekly data for all of 2012. For each week we record the % change in the USD/Euro, the % change in the US market, the % change in the British market, and the % change in the German market.



**3. You will now think of some real-life applications for statistical learning.**

(a) Describe three real-life applications in which classification might be useful. Describe the response, as well as the predictors. Is the goal of each application inference or prediction? Explain your answer.

(b) Describe three real-life applications in which regression might be useful. Describe the response, as well as the predictors. Is the goal of each application inference or prediction? Explain your answer.

(c) Describe three real-life applications in which cluster analysis might be useful.



**4. Describe the differences between a parametric and a non-parametric statistical learning approach. What are the advantages of a parametric approach to regression or classification (as opposed to a non-parametric approach)? What are its disadvantages?**



**5. This exercise involves the Auto data set studied in the lab. Make sure that the missing values have been removed from the data.**

```{r}
library(ISLR2)
data(Auto, package = 'ISLR2')
```


(a) Which of the predictors are quantitative, and which are qualitative?

(b) What is the range of each quantitative predictor? You can answer this using the range() function. (c) What is the mean and standard deviation of each quantitative predictor?

(d) Now remove the 10th through 85th observations. What is the range, mean, and standard deviation of each predictor in the subset of the data that remains?

(e) Using the full data set, investigate the predictors graphically, using scatterplots or other tools of your choice. Create some plots highlighting the relationships among the predictors. Comment on your findings.



**6. This exercise involves the Boston housing data set.**

```{r}
data(Boston, package = 'ISLR2')
```


(a) To begin, load in the Boston data set. The Boston data set is part of the ISLR2 library. How many rows are in this data set? How many columns? What do the rows and columns represent?

(b) Make some pairwise scatterplots of the predictors (columns) in this data set. Describe your findings.

(c) Are any of the predictors associated with per capita crime rate? If so, explain the relationship.

(d) Do any of the census tracts of Boston appear to have particularly high crime rates? Tax rates? Pupil-teacher ratios? Comment on the range of each predictor.

(e) How many of the census tracts in this data set bound the Charles river?

(f) What is the median pupil-teacher ratio among the towns in this data set?

(g) Which census tract of Boston has lowest median value of owner-occupied homes? What are the values of the other predictors for that census tract, and how do those values compare to the overall ranges for those predictors? Comment on your findings.

(h) In this data set, how many of the census tracts average more than seven rooms per dwelling? More than eight rooms per dwelling? Comment on the census tracts that average more than eight rooms per dwelling.
